The Proofs

### 2024-07-31 paper/tex/01-introduction.tex
fix capitalization

\paragraph*{Declarative Type System Specification (\cref{sec:overview,sec:declarative-system})} 
    We present \fexists---a second-order polymorphic type system that supports first-class
    existential and universal quantifiers, both of which can be instantiated
    impredicatively. To evade the undecidability results surrounding subtyping
    for \systemf, we base the system on \CBPV \cite{levy2006:cbpv}: we
    use \emph{polarization} of types and terms to formulate the restrictions
    that make typing decidable.

@@

\paragraph*{Type Inference Algorithm (\cref{sec:algorithm})} We give a type
  inference algorithm implementing the declarative specification. Because our
  system supports both existential and universal quantifiers, our implementation
  of the subtyping relation requires the use of \emph{anti-unification}---a dual
  to unification that finds the most specific generalization rather than the
  most general unifier. To our knowledge, anti-unification has never been used
  in the context of polymorphic type inference.

>>>>

\paragraph*{Declarative type system specification (\cref{sec:overview,sec:declarative-system})} 
    We present \fexists---a second-order polymorphic type system that supports first-class
    existential and universal quantifiers, both of which can be instantiated
    impredicatively. To evade the undecidability results surrounding subtyping
    for \systemf, we base the system on \CBPV \cite{levy2006:cbpv}: we
    use \emph{polarization} of types and terms to formulate the restrictions
    that make typing decidable.

@@

\paragraph*{Type inference algorithm (\cref{sec:algorithm})} 
  We give a type inference algorithm implementing the declarative specification.
  Because our system supports both existential and universal quantifiers, our
  implementation of the subtyping relation requires the use of
  \emph{anti-unification}---a dual to unification that finds the most specific
  generalization rather than the most general unifier. To our knowledge,
  anti-unification has never been used in the context of polymorphic type
  inference.

### 2024-07-31 paper/tex/02-overview.tex
fix capitalization

\paragraph*{Application of a Polymorphic Function}
  A polymorphic function can be applied to a value of a concrete type. In this
  case, the explicit type application is not required: the inference algorithm
  automatically instantiates the quantified variables. For instance, the
  polymorphic $[[len]]$ can be applied to a list of any positive type, and the
  resulting type will be inferred by the algorithm:
  $$[[· ; Φ ⊢ let x = len([one,two,three]); return x : ↑Int]]$$

\paragraph*{Polymorphic Self-application}
  Unlike predicative systems, \fexists permits self-application of polymorphic
  functions. For instance, $[[id]]$ can be applied to itself resulting in a
  suspended polymorphic computation of type $[[↓(∀α⁺.α⁺ → ↑α⁺)]]$, which, in
  turn, can be applied to itself:
  $$[[· ; Φ ⊢ let x = id(id); let y = x(x); return y : ↑↓(∀α⁺.α⁺ → ↑α⁺)]]$$

\paragraph*{Non-example: Polymorphic Arguments}
The focus of our system is the \emph{local} type inference
\cite{pierce2000:local}, which has certain limitations. In particular, the
instantiation does not happen when the polymorphic term is in the argument
position. For example, one could expect the following to be inferrable:
\mbox{$[[· ; Φ ⊢ let x = map(id, [two, three, nine]); return x : ↑ List Int]]$}.
However, to infer this, $[[id]]$ must be instantiated to \mbox{$[[Int → ↑Int]]$}
which requires the information to be propagated from the neighboring branch of
the syntax tree (\mbox{$[[ [two, three, nine] ]]$}), \ie bypass the locality.

@@

\paragraph*{Inferring the Common Supertype}
  The function $[[choose]]$ takes two arguments of the same
  polymorphic type. To apply it to two values of different (concrete) types, 
  the inference algorithm must find the minimal type they can be coerced 
  to---the least common supertype.

@@

\paragraph*{Syntactic Conventions}
We always consider terms and types up to alpha-equivalence, 
and substitution is capture-avoiding. In addition, we assume the quantification is associative,
for example, $[[∀pas.∀pbs.iN]]$ and $[[∀pas,pbs.iN]]$
represent the same type.

\paragraph*{Type Context and Type Well-formedness}

@@

\paragraph*{Least Upper Bound}
Eventually, the subtyping algorithm reduces the initial subtyping problem to a set of
constraints that need to be resolved. At this moment, another problem arises:
as one variable can occur in multiple constraints, the resolution becomes non-trivial.
In particular, to resolve the conjunction of $[[α̂⁺ ≥ iP]]$ and $[[α̂⁺ ≥ iQ]]$,
we would need to find the least upper bound of $[[iP]]$ and $[[iQ]]$.

>>>>

\paragraph*{Application of a polymorphic function}
  A polymorphic function can be applied to a value of a concrete type. In this
  case, the explicit type application is not required: the inference algorithm
  automatically instantiates the quantified variables. For instance, the
  polymorphic $[[len]]$ can be applied to a list of any positive type, and the
  resulting type will be inferred by the algorithm:
  $$[[· ; Φ ⊢ let x = len([one,two,three]); return x : ↑Int]]$$

\paragraph*{Polymorphic self-application}
  Unlike predicative systems, \fexists permits self-application of polymorphic
  functions. For instance, $[[id]]$ can be applied to itself resulting in a
  suspended polymorphic computation of type $[[↓(∀α⁺.α⁺ → ↑α⁺)]]$, which, in
  turn, can be applied to itself:
  $$[[· ; Φ ⊢ let x = id(id); let y = x(x); return y : ↑↓(∀α⁺.α⁺ → ↑α⁺)]]$$

\paragraph*{Non-example: polymorphic arguments}
The focus of our system is the \emph{local} type inference
\cite{pierce2000:local}, which has certain limitations. In particular, the
instantiation does not happen when the polymorphic term is in the argument
position. For example, one could expect the following to be inferrable:
\mbox{$[[· ; Φ ⊢ let x = map(id, [two, three, nine]); return x : ↑ List Int]]$}.
However, to infer this, $[[id]]$ must be instantiated to \mbox{$[[Int → ↑Int]]$}
which requires the information to be propagated from the neighboring branch of
the syntax tree (\mbox{$[[ [two, three, nine] ]]$}), \ie bypass the locality.

@@

\paragraph*{Inferring the common supertype}
  The function $[[choose]]$ takes two arguments of the same
  polymorphic type. To apply it to two values of different (concrete) types, 
  the inference algorithm must find the minimal type they can be coerced 
  to---the least common supertype.

@@

\paragraph*{Syntactic conventions}
We always consider terms and types up to alpha-equivalence, 
and substitution is capture-avoiding. In addition, we assume the quantification is associative,
for example, $[[∀pas.∀pbs.iN]]$ and $[[∀pas,pbs.iN]]$
represent the same type.

\paragraph*{Type context and type well-formedness}

@@

\paragraph*{Least upper bound}
Eventually, the subtyping algorithm reduces the initial subtyping problem to a set of
constraints that need to be resolved. At this moment, another problem arises:
as one variable can occur in multiple constraints, the resolution becomes non-trivial.
In particular, to resolve the conjunction of $[[α̂⁺ ≥ iP]]$ and $[[α̂⁺ ≥ iQ]]$,
we would need to find the least upper bound of $[[iP]]$ and $[[iQ]]$.

### 2024-07-31 paper/tex/03-declarative-system.tex
prettify + minor edits
fix capitalization

\paragraph*{Invariant Shifts}
As mentioned above, an important restriction that we put on the subtyping system
is that subtyping of shifted types requires their equivalence, as shown in
\ruleref{\ottdruleDOneShiftDLabel} and \ruleref{\ottdruleDOneShiftULabel}. The
reason for this is that if both of these rules were relaxed to the covariant
form, the subtyping relation would become equivalent to the standard subtyping
of \systemf, which is undecidable \cite{tiuryn-urzczyn-96}. 
Relaxations of this condition are discussed in \cref{sec:weakening-invariant}.

@@

  \vspace{\baselineskip}
% \subsubsection{Properties of Declarative Typing}
  An important property that the declarative system has is
  that the declarative specification is correctly defined for
  equivalence classes.

@@

\paragraph*{Type-level Translation}

@@

\paragraph*{Term-level Translation}

@@

\begin{table}[h]
\centering
\begin{tabular}{rll}
\hline
\multicolumn{2}{c}{\textbf{Elaboration Judgment}} & \textbf{The Soundness Property} \\
\hline
Negative Subtyping & $[[Γ ⊢ iN ≤ iM ⤳ ft]]$ & $[[|Γ| ; · ⊢ ft : |iN| → |iM| ]]$ \\
Positive Subtyping & $[[Γ ⊢ iP ≥ iQ ⤳ ft]]$ & $[[|Γ| ; · ⊢ ft : |iQ| → |iP| ]]$ \\
Negative Typing & $[[Γ; Φ ⊢ c : iN ⤳ ft]]$ & $[[|Γ| ; |Φ| ⊢ ft : |iN|]]$ \\
Positive Typing & $[[Γ; Φ ⊢ v : iP ⤳ ft]]$ & $[[|Γ| ; |Φ| ⊢ ft : |iP|]]$ \\
Application Typing & $[[Γ ; Φ ⊢ iN ● args ⇒> iM ⤳ fe; fargs]]$ & $[[|Γ| ; |Φ|, x:|iN| ⊢ fe (x ● fargs) : |iM|]]$ \\
\hline
\systemf Typing & $[[fΓ ; fΦ ⊢ ft : fT ⤳ c]]$ & $[[+|fΓ|+ ; +|fΦ|+ ⊢ c : ↑+|fT|+]]$ \\
\hline
\end{tabular}
\caption{Soundness of the Elaboration:
  the elaboration judgment on the left implies the typing judgment on the right}
  \label{tab:elaboration-soundness}
\end{table}

>>>>

\paragraph*{Invariant shifts}
As mentioned above, an important restriction that we put on the subtyping system
is that subtyping of shifted types requires their equivalence, as shown in
\ruleref{\ottdruleDOneShiftDLabel} and \ruleref{\ottdruleDOneShiftULabel}. The
reason for this is that if both of these rules were relaxed to the covariant
form, the subtyping relation would become equivalent to the standard subtyping
of \systemf, which is undecidable \cite{tiuryn-urzczyn-96}. 
Relaxations of this condition are discussed in \cref{sec:weakening-invariant}.

@@

  % \vspace{\baselineskip}
% \subsubsection{Properties of Declarative Typing}
  An important property that the declarative system has is
  that the declarative specification is correctly defined for
  equivalence classes.

@@

\paragraph*{Type-level translation}

@@

\paragraph*{Term-level translation}

@@

\begin{figure}[t]
\centering
\begin{tabular}{rll}
\hline
\multicolumn{2}{c}{\textbf{Elaboration Judgment}} & \textbf{The Soundness Property} \\
\hline
Negative Subtyping & $[[Γ ⊢ iN ≤ iM ⤳ ft]]$ & $[[|Γ| ; · ⊢ ft : |iN| → |iM| ]]$ \\
Positive Subtyping & $[[Γ ⊢ iP ≥ iQ ⤳ ft]]$ & $[[|Γ| ; · ⊢ ft : |iQ| → |iP| ]]$ \\
Negative Typing & $[[Γ; Φ ⊢ c : iN ⤳ ft]]$ & $[[|Γ| ; |Φ| ⊢ ft : |iN|]]$ \\
Positive Typing & $[[Γ; Φ ⊢ v : iP ⤳ ft]]$ & $[[|Γ| ; |Φ| ⊢ ft : |iP|]]$ \\
Application Typing & $[[Γ ; Φ ⊢ iN ● args ⇒> iM ⤳ fe; fargs]]$ & $[[|Γ| ; |Φ|, x:|iN| ⊢ fe (x ● fargs) : |iM|]]$ \\
\hline
\systemf Typing & $[[fΓ ; fΦ ⊢ ft : fT ⤳ c]]$ & $[[+|fΓ|+ ; +|fΦ|+ ⊢ c : ↑+|fT|+]]$ \\
\hline
\end{tabular}
\caption{Soundness of the Elaboration:
  the elaboration judgment on the left implies the typing judgment on the right}
  \label{tab:elaboration-soundness}
\end{figure}

### 2024-07-31 paper/tex/04-algorithm.tex
fix capitalization

\paragraph*{Algorithmic Variables}
To design a mode-correct inference system, we slightly modify the language we operate on.
The entities (terms, types, contexts) that the algorithm manipulates we call \emph{algorithmic}. 
They extend the previously defined declarative terms and types by adding 
\emph{algorithmic type variables} (\aka unification variables). 
The algorithmic variables represent unknown types, 
which cannot be inferred immediately but are promised to be instantiated
as the algorithm proceeds.

@@

\paragraph*{Algorithmic Types}
The syntax of algorithmic types extends the declarative syntax by adding
algorithmic variables as new terminals. We add positive algorithmic variables $[[α̂⁺]]$ 
to the syntax of positive types, and negative algorithmic variables $[[α̂⁻]]$ to the 
syntax of negative types. All the constructors of the system can be applied 
to \emph{algorithmic} types, however, algorithmic variables cannot be abstracted by the
quantifiers $[[∀]]$ and $[[∃]]$.

\paragraph*{Algorithmic Contexts $[[Ξ]]$ and Well-formedness}
To specify when algorithmic types are well-formed, we define algorithmic
contexts $[[Ξ]]$ as sets of algorithmic variables. Then
$[[Γ ; Ξ ⊢ uP]]$ and $[[Γ ; Ξ ⊢ uN]]$ represent the well-formedness judgment of
algorithmic terms defined as expected. Informally, they check that all free
declarative variables are in $[[Γ]]$, and all free algorithmic variables are in
$[[Ξ]]$. Most of the rules are inherited from the well-formedness of
\emph{declarative} types: the declarative variables are checked to belong to the
context $[[Γ]]$, the quantifiers extend the context $[[Γ]]$, type constructors are
well-formed congruently. As we extend the syntax with algorithmic variables, we
also add two base-case rules for them: \ruleref{\ottdruleWFATPUVarLabel} and
\ruleref{\ottdruleWFATNUVarLabel} (see \cref{fig:algo-wf}).

@@

\paragraph*{Instantiation Context $[[Θ]]$}
When one instantiates an algorithmic variable, one may only use type variables
\emph{available in its scope}. As such, each algorithmic variable must remember
the context at the moment when it was introduced. In our algorithm, this
information is represented by an \emph{instantiation context} $[[Θ]]$---a set of
pairs associating algorithmic variables and declarative contexts.

\paragraph*{Algorithmic Substitution}
We define the algorithmic substitution $[[uσ]]$ as a mapping from algorithmic
variables to \emph{declarative} types. The signature $[[Θ ⊢ uσ : Ξ]]$
specifies the domain and the range of $[[uσ]]$: for each variable $[[α̂±]]$
in $[[Ξ]]$, there exists an corresponding entry in $[[Θ]]$ associating 
$[[α̂±]]$ with a declarative context $[[Γ]]$ such that $[[ [uσ]α̂± ]]$
is well-formed in $[[Γ]]$. In addition, we assume that $[[uσ]]$
acts as the identity on the variables not in $[[Ξ]]$.

\paragraph*{Algorithmic Normalization}
Similarly to well-formedness, the normalization of algorithmic types is defined
by extending the declarative definition (\cref{fig:type-nf}) with the
algorithmic variables. To the rules repeating the declarative normalization, we
add rules saying that normalization is trivial on algorithmic variables.

@@

\paragraph*{Auxiliary Functions}
\begin{itemize}
  \item
    We define $[[dom(SC)]]$---the domain of constraint set $[[SC]]$---as a set of
    algorithmic variables that it restricts. Similarly, we define $[[dom(Θ)]]$---the
    domain of instantiation context $[[Θ]]$---as a set of algorithmic variables that
    $[[Θ]]$ associates with their contexts.  
  \item We write $[[Θ(α̂±)]]$ to denote the
    declarative context associated with $[[α̂±]]$ in $[[Θ]]$. 
  \item $[[uv uN]]$ and $[[uv uP]]$ denote the set of free algorithmic variables in $[[uN]]$ and
    $[[uP]]$ respectively.
  \item We write $[[Γ ⊢ Θ]]$ to denote that each
    declarative context associated with an algorithmic variable in $[[Θ]]$ is 
    a subcontext (subset) of $[[Γ]]$.
\end{itemize}

\paragraph*{Equivalent Substitutions}
In the proofs of the algorithm correctness, 
we often state or require that two substitutions are equivalent
on a given set of algorithmic variables.
We denote it as $[[Θ ⊢ uσ' ≈ uσ : Ξ]]$ meaning that
for each $[[α̂±]]$ in $[[Ξ]]$, substitutions $[[uσ]]$ and $[[uσ']]$
map $[[α̂±]]$ to types equivalent in the corresponding context:
$[[ Θ(α̂±) ⊢ [uσ']α̂± ≈ [uσ]α̂± ]]$.

@@

\paragraph*{The Type Upgrade}
The type upgrade algorithm uses the least upper bound algorithm as a subroutine.
It exploits the idea that the free variables of a positive type $[[iQ]]$
cannot disappear in its subtypes (see \cref{prop:subtyping-preserves-fv}). 
It means that if 
a type $[[iP]]$ has free variables not occurring 
in $[[iP']]$, then any common supertype of $[[iP]]$
and $[[iP']]$ must not contain these variables either.
This way, any supertype of $[[iP]]$
not containing certain variables $[[pnas]]$ must also be 
a supertype of $[[iP' = [pnbs/pnas]iP ]]$, where $[[pnbs]]$ are fresh;
and vice versa: any common supertype of $[[iP]]$ and $[[iP']]$
does not contain $[[pnas]]$ or $[[pnbs]]$.

@@

\paragraph*{The Least Upper Bound}
The Least Upper Bound algorithm we use operates on \emph{positive}
types. This way, the inference rules of the algorithm
analyze the three possible shapes of the input types:
a variable type, an existential type, and a shifted computation.

@@

\paragraph*{The Anti-Unification Rule}
  Rule \ruleref{\ottdruleAUAULabel} is the base case of the anti-unification
  algorithm. If the congruent rules are not applicable, 
  it means that the input types have a substantially different structure,
  and thus, the only option is to create a `hole'. 
  There are three important aspects of this rule that we would like to discuss.

@@

\paragraph*{Return and Thunk}
  Rules \ruleref{\ottdruleATThunkLabel} and \ruleref{\ottdruleATReturnLabel}
  are similar to the declarative rules: they make a recursive call
  to type the body of the thunk or the return expression and
  put the shift on top of the result.

@@

\paragraph*{Let Binders}
  Rule \ruleref{\ottdruleATVarLetLabel} represents the type inference of a
  standard let binder. It infers the type of the bound value $[[v]]$, and makes
  a recursive call to infer the type of the body in the extended context.

@@

\paragraph*{Application to an Empty List of Arguments}
  Rule \ruleref{\ottdruleATEmptyAppLabel}
  is the base case of application inference. 
  If the list of arguments is empty, 
  the inferred type is the type of the head,
  and the algorithm returns it after normalizing.

\paragraph*{Application of a Polymorphic Type $[[∀]]$}
  Rule \ruleref{\ottdruleATForallAppLabel},
  analogously to the declarative case,
  is the rule ensuring the implicit elimination of the universal quantifiers. 
  This is the place where the algorithmic variables are introduced.
  The algorithm simply replaces the quantified variables 
  $[[pas]]$ with fresh algorithmic variables $[[puas]]$,
  and makes a recursive call in the extended context.

@@

\paragraph*{Application of an Arrow Type}
  Rule \ruleref{\ottdruleATArrowAppLabel}
  is the main rule of algorithmic application inference.
  It is applied when the head has an arrow type $[[uQ → uN]]$.
  First, it infers the type of the first argument $[[v]]$,
  and then, calling the algorithmic subtyping,
  finds $[[SC1]]$---the \emph{minimal} constraint ensuring that 
  $[[uQ]]$ is a supertype of the type of $[[v]]$.
  Then it makes a recursive call applying $[[uN]]$ to the rest of the arguments 
  and merges the resulting constraint with $[[SC1]]$.

>>>>

\paragraph*{Algorithmic variables}
To design a mode-correct inference system, we slightly modify the language we operate on.
The entities (terms, types, contexts) that the algorithm manipulates we call \emph{algorithmic}. 
They extend the previously defined declarative terms and types by adding 
\emph{algorithmic type variables} (\aka unification variables). 
The algorithmic variables represent unknown types, 
which cannot be inferred immediately but are promised to be instantiated
as the algorithm proceeds.

@@

\paragraph*{Algorithmic types}
The syntax of algorithmic types extends the declarative syntax by adding
algorithmic variables as new terminals. We add positive algorithmic variables $[[α̂⁺]]$ 
to the syntax of positive types, and negative algorithmic variables $[[α̂⁻]]$ to the 
syntax of negative types. All the constructors of the system can be applied 
to \emph{algorithmic} types, however, algorithmic variables cannot be abstracted by the
quantifiers $[[∀]]$ and $[[∃]]$.

\paragraph*{Algorithmic contexts $[[Ξ]]$ and well-formedness}
To specify when algorithmic types are well-formed, we define algorithmic
contexts $[[Ξ]]$ as sets of algorithmic variables. Then
$[[Γ ; Ξ ⊢ uP]]$ and $[[Γ ; Ξ ⊢ uN]]$ represent the well-formedness judgment of
algorithmic terms defined as expected. Informally, they check that all free
declarative variables are in $[[Γ]]$, and all free algorithmic variables are in
$[[Ξ]]$. Most of the rules are inherited from the well-formedness of
\emph{declarative} types: the declarative variables are checked to belong to the
context $[[Γ]]$, the quantifiers extend the context $[[Γ]]$, type constructors are
well-formed congruently. As we extend the syntax with algorithmic variables, we
also add two base-case rules for them: \ruleref{\ottdruleWFATPUVarLabel} and
\ruleref{\ottdruleWFATNUVarLabel} (see \cref{fig:algo-wf}).

@@

\paragraph*{Instantiation context $[[Θ]]$}
When one instantiates an algorithmic variable, one may only use type variables
\emph{available in its scope}. As such, each algorithmic variable must remember
the context at the moment when it was introduced. In our algorithm, this
information is represented by an \emph{instantiation context} $[[Θ]]$---a set of
pairs associating algorithmic variables and declarative contexts.

\paragraph*{Algorithmic substitution}
We define the algorithmic substitution $[[uσ]]$ as a mapping from algorithmic
variables to \emph{declarative} types. The signature $[[Θ ⊢ uσ : Ξ]]$
specifies the domain and the range of $[[uσ]]$: for each variable $[[α̂±]]$
in $[[Ξ]]$, there exists an corresponding entry in $[[Θ]]$ associating 
$[[α̂±]]$ with a declarative context $[[Γ]]$ such that $[[ [uσ]α̂± ]]$
is well-formed in $[[Γ]]$. In addition, we assume that $[[uσ]]$
acts as the identity on the variables not in $[[Ξ]]$.

\paragraph*{Algorithmic normalization}
Similarly to well-formedness, the normalization of algorithmic types is defined
by extending the declarative definition (\cref{fig:type-nf}) with the
algorithmic variables. To the rules repeating the declarative normalization, we
add rules saying that normalization is trivial on algorithmic variables.

@@

\paragraph*{Auxiliary functions}
\begin{itemize}
  \item
    We define $[[dom(SC)]]$---the domain of constraint set $[[SC]]$---as a set of
    algorithmic variables that it restricts. Similarly, we define $[[dom(Θ)]]$---the
    domain of instantiation context $[[Θ]]$---as a set of algorithmic variables that
    $[[Θ]]$ associates with their contexts.  
  \item We write $[[Θ(α̂±)]]$ to denote the
    declarative context associated with $[[α̂±]]$ in $[[Θ]]$. 
  \item $[[uv uN]]$ and $[[uv uP]]$ denote the set of free algorithmic variables in $[[uN]]$ and
    $[[uP]]$ respectively.
  \item We write $[[Γ ⊢ Θ]]$ to denote that each
    declarative context associated with an algorithmic variable in $[[Θ]]$ is 
    a subcontext (subset) of $[[Γ]]$.
\end{itemize}

\paragraph*{Equivalent substitutions}
In the proofs of the algorithm correctness, 
we often state or require that two substitutions are equivalent
on a given set of algorithmic variables.
We denote it as $[[Θ ⊢ uσ' ≈ uσ : Ξ]]$ meaning that
for each $[[α̂±]]$ in $[[Ξ]]$, substitutions $[[uσ]]$ and $[[uσ']]$
map $[[α̂±]]$ to types equivalent in the corresponding context:
$[[ Θ(α̂±) ⊢ [uσ']α̂± ≈ [uσ]α̂± ]]$.

@@

\paragraph*{Type upgrade}
The type upgrade algorithm uses the least upper bound algorithm as a subroutine.
It exploits the idea that the free variables of a positive type $[[iQ]]$
cannot disappear in its subtypes (see \cref{prop:subtyping-preserves-fv}). 
It means that if 
a type $[[iP]]$ has free variables not occurring 
in $[[iP']]$, then any common supertype of $[[iP]]$
and $[[iP']]$ must not contain these variables either.
This way, any supertype of $[[iP]]$
not containing certain variables $[[pnas]]$ must also be 
a supertype of $[[iP' = [pnbs/pnas]iP ]]$, where $[[pnbs]]$ are fresh;
and vice versa: any common supertype of $[[iP]]$ and $[[iP']]$
does not contain $[[pnas]]$ or $[[pnbs]]$.

@@

\paragraph*{The least upper bound}
The Least Upper Bound algorithm we use operates on \emph{positive}
types. This way, the inference rules of the algorithm
analyze the three possible shapes of the input types:
a variable type, an existential type, and a shifted computation.

@@

\paragraph*{The anti-unification rule}
  Rule \ruleref{\ottdruleAUAULabel} is the base case of the anti-unification
  algorithm. If the congruent rules are not applicable, 
  it means that the input types have a substantially different structure,
  and thus, the only option is to create a `hole'. 
  There are three important aspects of this rule that we would like to discuss.

@@

\paragraph*{Return and thunk}
  Rules \ruleref{\ottdruleATThunkLabel} and \ruleref{\ottdruleATReturnLabel}
  are similar to the declarative rules: they make a recursive call
  to type the body of the thunk or the return expression and
  put the shift on top of the result.

@@

\paragraph*{Let binders}
  Rule \ruleref{\ottdruleATVarLetLabel} represents the type inference of a
  standard let binder. It infers the type of the bound value $[[v]]$, and makes
  a recursive call to infer the type of the body in the extended context.

@@

\paragraph*{Application to an empty list of arguments}
  Rule \ruleref{\ottdruleATEmptyAppLabel}
  is the base case of application inference. 
  If the list of arguments is empty, 
  the inferred type is the type of the head,
  and the algorithm returns it after normalizing.

\paragraph*{Application of a polymorphic type $[[∀]]$}
  Rule \ruleref{\ottdruleATForallAppLabel},
  analogously to the declarative case,
  is the rule ensuring the implicit elimination of the universal quantifiers. 
  This is the place where the algorithmic variables are introduced.
  The algorithm simply replaces the quantified variables 
  $[[pas]]$ with fresh algorithmic variables $[[puas]]$,
  and makes a recursive call in the extended context.

@@

\paragraph*{Application of an arrow type}
  Rule \ruleref{\ottdruleATArrowAppLabel}
  is the main rule of algorithmic application inference.
  It is applied when the head has an arrow type $[[uQ → uN]]$.
  First, it infers the type of the first argument $[[v]]$,
  and then, calling the algorithmic subtyping,
  finds $[[SC1]]$---the \emph{minimal} constraint ensuring that 
  $[[uQ]]$ is a supertype of the type of $[[v]]$.
  Then it makes a recursive call applying $[[uN]]$ to the rest of the arguments 
  and merges the resulting constraint with $[[SC1]]$.

### 2024-07-31 paper/tex/05-proof.tex
fix capitalization

\paragraph*{Mutuality of the Soundness and Completeness Proofs}

@@

\paragraph*{Inductive Metric}
    The soundness and completeness lemmas are proven by \emph{mutual} induction. 
    Since the declarative and the algorithmic systems do not depend on each other,
    we must introduce a uniform \emph{metric} on which the induction is conducted.
    We define the metric gradually, starting from the
    auxiliary function---the size of a judgment $\size{J}$.

>>>>

\paragraph*{Mutuality of the soundness and completeness proofs}

@@

\paragraph*{Inductive metric}
    The soundness and completeness lemmas are proven by \emph{mutual} induction. 
    Since the declarative and the algorithmic systems do not depend on each other,
    we must introduce a uniform \emph{metric} on which the induction is conducted.
    We define the metric gradually, starting from the
    auxiliary function---the size of a judgment $\size{J}$.

### 2024-07-31 paper/tex/06-extensions.tex
fix capitalization

\vspace{\baselineskip}
\paragraph*{Solution 1: Declarative Normalization}
One way to resolve this ambiguity is to fix the order of quantifiers.  
The algorithm already performs the ordering of the quantifiers in the normalization procedure.
This way, we could require the inferred type to be normalized
to specify the order of the quantifiers:
$$\ottdruleDTTypeAppOrderedLabeled{}$$

@@

\vspace{\baselineskip}
\paragraph*{Solution 2: Elementary Type Inference}
An alternative approach to explicit type application
 was proposed by \cite{zhao22:elementary}.
In this work, the subtyping relation is restricted 
in such a way that 
$[[∀α⁺.∀β⁺.iN]]$ and $[[∀β⁺.∀α⁺.iN]]$ are \emph{not} mutual subtypes 
(as long as $[[α⁺ ∊ fv(iN)]]$ and $[[β⁺ ∊ fv(iN)]]$).
It implies that the order of the quantifiers of the inferred type is unique,
and thus, the explicit type application is unambiguous.

@@

\vspace{\baselineskip}
\paragraph*{The Algorithm}
    The algorithmic system is bidirectionalized in a similar way. 
    Each typing premise and the conclusion of the rule are oriented
    to either `checking` ($[[⇐]]$) or `synthesizing` ($[[⇒]]$) mode,
    with respect to the declarative system.

>>>>

\vspace{\baselineskip}
\paragraph*{Solution 1: declarative normalization}
One way to resolve this ambiguity is to fix the order of quantifiers.  
The algorithm already performs the ordering of the quantifiers in the normalization procedure.
This way, we could require the inferred type to be normalized
to specify the order of the quantifiers:
$$\ottdruleDTTypeAppOrderedLabeled{}$$

@@

\vspace{\baselineskip}
\paragraph*{Solution 2: elementary type inference}
An alternative approach to explicit type application
 was proposed by \cite{zhao22:elementary}.
In this work, the subtyping relation is restricted 
in such a way that 
$[[∀α⁺.∀β⁺.iN]]$ and $[[∀β⁺.∀α⁺.iN]]$ are \emph{not} mutual subtypes 
(as long as $[[α⁺ ∊ fv(iN)]]$ and $[[β⁺ ∊ fv(iN)]]$).
It implies that the order of the quantifiers of the inferred type is unique,
and thus, the explicit type application is unambiguous.

@@

\vspace{\baselineskip}
\paragraph*{The algorithm}
    The algorithmic system is bidirectionalized in a similar way. 
    Each typing premise and the conclusion of the rule are oriented
    to either `checking` ($[[⇐]]$) or `synthesizing` ($[[⇒]]$) mode,
    with respect to the declarative system.

### 2024-07-31 proofs/tex/decl-typing-lemmas.tex
fix capitalization

\begin{definition}[Number of prenex quantifiers]
    Let us define $\npq{[[iN]]}$ and $\npq{[[iP]]}$ as the number of prenex quantifiers in these types, i.e.
    \begin{itemize}
        \item [$+$] $\npq{[[∃nas.iP]]} = |[[nas]]|$, if $[[iP ≠ ∃nbs.iP']]$,
        \item [$-$] $\npq{[[∀pas.iN]]} = |[[pas]]|$, if $[[iN ≠ ∀pbs.iN']]$.
    \end{itemize}
\end{definition}

>>>>

\begin{definition}[Number of Prenex Quantifiers]
    Let us define $\npq{[[iN]]}$ and $\npq{[[iP]]}$ as the number of prenex quantifiers in these types, i.e.
    \begin{itemize}
        \item [$+$] $\npq{[[∃nas.iP]]} = |[[nas]]|$, if $[[iP ≠ ∃nbs.iP']]$,
        \item [$-$] $\npq{[[∀pas.iN]]} = |[[pas]]|$, if $[[iN ≠ ∀pbs.iN']]$.
    \end{itemize}
\end{definition}

### 2024-07-31 proofs/tex/ord-lemmas.tex
fix capitalization

\begin{definition} [Collision-free bijection]
  We say that a bijection $\mu : A \leftrightarrow B$ between sets of
  variables is \textbf{collision-free on sets} $P$ and $Q$ if and only if
  \begin{enumerate}
    \item $\mu(P \cap A) \cap Q = \emptyset$
    \item $\mu(Q \cap A) \cap P = \emptyset$
  \end{enumerate}
\end{definition}

>>>>

\begin{definition} [Collision-free Bijection]
  We say that a bijection $\mu : A \leftrightarrow B$ between sets of
  variables is \textbf{collision-free on sets} $P$ and $Q$ if and only if
  \begin{enumerate}
    \item $\mu(P \cap A) \cap Q = \emptyset$
    \item $\mu(Q \cap A) \cap P = \emptyset$
  \end{enumerate}
\end{definition}
